---
title: "RAW plankton loader tool"
---

This tool reads raw plankton data from national datasets which has already been aggregated into a single dataframe. This data is combined with the CPR data from the previous step and lifeforms are identified from a Master List before saving the output.

Author: Matthew Holland
Contact: matt.holland@plymouth.ac.uk
Latest version date: 1 March 2022
```{r}
#clear R environment
rm(list = ls()) 

#turn off scientific notation
options(scipen=999)

#enter the directory for where the processed data is to be stored
dir_main <- "../Data_processed/"

#enter the directory for where the raw data is stored
dir_raw <- "../Data_raw/"
```
Load required packages and install if not already installed
```{r, include=FALSE}
#check if all required packages are installed. Install them if they are not present. 
#Then load required packages
list.of.packages <- c("tidyverse", "data.table", "readxl", "janitor", "fst")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
rm(list.of.packages)
```
Create subdirectories for processed data if not already done
```{r}
#generate directory for processed data
dir.create(file.path(dir_main), showWarnings = FALSE)
```
Read in the DASSH Master List for matching to the raw data
```{r}
df_master <- read_excel(paste0(dir_raw, "Masterlist-V5_working.xlsx"), sheet = "SpeciesInfo")
df_master <- janitor::clean_names(df_master) #make the column names more R friendly
```
Read in the RAW data
```{r}
#load the plankton abundance data
df_abund <- read_fst(paste0(dir_raw, "df_abund.fst")) %>%
  filter(abundance > 0 & !is.na(aphia_id))
```
Process the RAW data to have clean scientific names and Aphia IDs 
```{r}
#subset to remove samples deeper than a user-specified cut-off depth
cutoff <- 60
df_abund <- subset(df_abund, is.na(depth_max) | depth_max <= cutoff)

#function to match data from WoRMS to the abundance datasets
load_worms <- function(x) {
  unique_aphias <- x %>%
    dplyr::select(aphia_id) %>%
    unique() %>%
    dplyr::mutate(aphia_id = as.integer(aphia_id))

  
  if(file.exists(paste0(dir_main, "aphia_ids_and_names_from_worms.csv"))){
    #load the file if it exists
    df_wms <- fread(paste0(dir_main, "aphia_ids_and_names_from_worms.csv"))
    df_wms <- data.frame(lapply(df_wms, as.character), stringsAsFactors=FALSE)
  } else {
    #otherwise request data from WoRMS
    df_wms <- worms::wormsbyid(unique_aphias$aphia_id)
    df_wms <- data.frame(lapply(df_wms, as.character), stringsAsFactors=FALSE)
  }
  
  #check if all Aphia IDs in the dataset are present. If not, download the data again to update it.
  if(!all(unique(x$aphia_id) %in% df_wms$AphiaID)){
      
      #find the additional missing IDs
      missing_ids <- sort(unique(x$aphia_id))[!sort(unique(x$aphia_id)) %in% df_wms$AphiaID]
      
      unique_aphias <- unique_aphias %>%
      filter(aphia_id %in% all_of(missing_ids))
     
      df_wms2 <- worms::wormsbyid(unique_aphias$aphia_id)
      df_wms2 <- data.frame(lapply(df_wms2, as.character), stringsAsFactors=FALSE)
  
      df_wms <- rbind(df_wms, df_wms2)
      fwrite(df_wms, paste0(dir_main, "aphia_ids_and_names_from_worms.csv"))
}

  return(df_wms)

}

df_wms <- load_worms(x=df_abund)

#save the updated Aphia ID and taxon list
fwrite(df_wms, paste0(dir_main, "aphia_ids_and_names_from_worms.csv"))

#replace taxon scientific names from WoRMS for relevant Aphia IDs in abundance data and sum per sample
df_abund$taxon2 <- df_wms$scientificname[match(df_abund$aphia_id, df_wms$AphiaID)]
df_abund$taxon2[is.na(df_abund$taxon2)] <- df_abund$taxon[is.na(df_abund$taxon2)]
df_abund$taxon <- df_abund$taxon2
df_abund$taxon2 <- NULL
rm(df_wms)

df_abund <- df_abund %>%
  dplyr::select(-comment) %>%
  group_by(across(c(-abundance))) %>%
  dplyr::summarise(abundance = sum(abundance),
            .groups = "drop")
```
Convert units to be consistent across datasets
```{r, warning=FALSE}
#recode the abundance units to be consistent across datasets
df_abund$abundance_type_units <- tolower(df_abund$abundance_type_units)

df_abund$abundance_type_units <- recode_factor(df_abund$abundance_type_units, 
                                               "cell numbers per litre of water" = "cells/l",
                                               "cells/litre" = "cells/l",
                                               "hose sample cells/l" = "cells/l",
                                               "cells / l" = "cells/l",
                                               "numbers of individuals per cubic metre" = "cells/m3",
                                               "individuals/m3" = "cells/m3",
                                               "number/m3 (number individuos per cubic meter)" = "cells/m3",
                                               "numbers / m3" = "cells/m3",
                                               "abundance m-3" = "cells/m3",
                                               "no of individuals per m3" = "cells/m3",
                                               "count m-3" = "cells/m3",
                                               "indiv. / m3" = "cells/m3",
                                               "net sample (wp2) ind/m3" = "cells/m3",
                                               "nr/m3" = "cells/m3",
                                               "cells per ml" = "cells/ml",
                                               "count" = "cells/m3")
unique(df_abund$abundance_type_units)

#convert abundance units to a consistent cells/l
df_abund$abundance <- ifelse(df_abund$abundance_type_units=="cells/m3",
       df_abund$abundance,
       ifelse(df_abund$abundance_type_units=="cells/ml",
              df_abund$abundance*1000*1000,
              ifelse(df_abund$abundance_type_units=="cells/l",
                     df_abund$abundance*1000,
                     df_abund$abundance)))
```
Get the data into the same format as the CPR so it can be combined
```{r}
#work out a short dataset ID
df_abund <- df_abund %>%
  dplyr::mutate(cp=sub('-.*', '', contracting_party),
         inst=sub('\\).*', '', sub('.*\\(', '', contracting_party))) %>%
  arrange(cp, inst) %>%
  group_by(cp, inst) %>%
  dplyr::mutate(dset=as.numeric(factor(dataset_name))) %>%
  ungroup() %>%
  dplyr::mutate(data_id=paste(cp, inst, dset, sep="-"))

#save the dataset indices
df_datasets <- df_abund %>%
  dplyr::select(cp, inst, dset, data_id, contracting_party, dataset_name) %>%
  distinct() %>%
  bind_rows(data.frame(cp="UK",
                       inst="MBA",
                       dset=1,
                       data_id="UK-MBA-1",
                       contracting_party="UK-Marine Biologicical Association (MBA)",
                       dataset_name="CPR dataset")) %>%
  arrange(data_id)

#save the datasets info dataframe
fwrite(df_datasets, file=paste0(dir_main, "dataset_ids", ".csv"))

df_abund <- df_abund %>%
  dplyr::select(-cp, -inst, -dset)
```
Convert the UK-SAMS data to point (station) rather than distributed so that it is not excluded from the analysis
```{r, warning=FALSE, include=FALSE}
if(any(grepl("SAMS", unique(df_abund$data_id)))){
  
  df_abund_temp <- df_abund %>%
    filter(!grepl("SAMS", data_id))
  
  df_abund <- df_abund %>%
    filter(grepl("SAMS", data_id)) %>%
    group_by(latitude, longitude, date, depth_min, depth_max) %>%
    dplyr::mutate(hour = as.character(sample(23,1, replace = FALSE)),
           minute = as.character(sample(59,1, replace = FALSE)),) %>%
    ungroup() %>%
    dplyr::mutate(minute = ifelse(nchar(minute)<2, paste0("0", minute), minute)) %>%
    dplyr::mutate(time = paste(hour, minute, "00", sep=":")) %>%
    dplyr::select(-c(hour, minute)) %>%
    dplyr::mutate(longitude = -5.5,
           latitude = 56.5) %>%
    bind_rows(df_abund_temp)
  
  rm(df_abund_temp)
    
}
```
Split time and date
```{r, warning=FALSE}
#split the date variable into year, month, day
#format date column
df_abund$year <- format(df_abund$date, format="%Y")
df_abund$month <- format(df_abund$date, format="%m")
df_abund$day <- format(df_abund$date, format="%d")

#split the time variable into hour, minute
#format time column
df_abund$time <- ifelse(df_abund$time=="", NA, df_abund$time)
df_abund$hour <- sub(':.*', '', df_abund$time)
df_abund$minute <- sub(".*?:",'',df_abund$time)
df_abund$minute <- sub(':.*', '', df_abund$minute)

#rename variables and order according to the process used for CPR data
df_abund <- df_abund %>%
  dplyr::rename("lon"=longitude,
         "lat"=latitude,
         "count"=abundance) %>%
  dplyr::select(data_id,
         year, month, day, hour, minute, time, lat, lon, size_class, taxon, aphia_id, count)
```
Load the CPR data and combine with df_abund
```{r}
df_cpr <- read_fst(path=paste0(dir_main, "df_cpr.fst")) %>%
  dplyr::mutate(size_class = NA,
         .before="taxon")

df_abund <- df_abund %>%
  dplyr::select(names(df_cpr))

df_abund <- rbind(df_cpr, df_abund)

rm(df_cpr)

df_abund <- as.data.table(df_abund)
```
Match the Master List classifications to the plankton data
```{r}
#dataframe containing the lifeform trait criteria
lf_select <- rbind(data.frame(column1 = "plankton_type", cat1="Phytoplankton",
                           column2="phytoplankton_type", cat2="Diatom",
                           name="diatom"),
data.frame(column1 = "plankton_type", cat1="Phytoplankton",
                           column2="phytoplankton_type", cat2="Dinoflagellate",
                           name="dinoflagellate"),
data.frame(column1 = "phytoplankton_type", cat1="Diatom",
                           column2="phyto_depth", cat2="Pelagic",
                           name="pelagic_diatoms"),
data.frame(column1 = "phytoplankton_type", cat1="Diatom",
                           column2="phyto_depth", cat2="Tychopelagic",
                           name="tycho_diatoms"),
data.frame(column1 = "plankton_type", cat1="Phytoplankton",
                           column2="phytoplankton_size", cat2="Lg",
                           name="lg_phyto"),
data.frame(column1 = "plankton_type", cat1="Phytoplankton",
                           column2="phytoplankton_size", cat2="Sm",
                           name="sm_phyto"),
data.frame(column1 = "copepod", cat1="Y",
                           column2="cop_size", cat2="Lg",
                           name="lg_copepods"),
data.frame(column1 = "copepod", cat1="Y",
                           column2="cop_size", cat2="Sm",
                           name="sm_copepods"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="zoo_habitat", cat2="Holoplankton",
                           name="holoplankton"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="zoo_habitat", cat2="Meroplankton",
                           name="meroplankton"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="zoo_type", cat2="Fish",
                           name="fishlarvae"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="zoo_diet", cat2=c("Herbivore", "Omnivore", "Ambiguous"),
                           name="noncarniv"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="zoo_diet", cat2="Carnivore",
                           name="carniv"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="zoo_type", cat2="Crustacean",
                           name="crustacean"),
data.frame(column1 = "plankton_type", cat1="Phytoplankton",
                           column2="plankton_type", cat2="Phytoplankton",
                           name="phytoplankton"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="copepod", cat2="Y",
                           name="copepods"),
data.frame(column1 = "plankton_type", cat1="Zooplankton",
                           column2="gelatinous", cat2="Y",
                           name="gelatinous")
)

#function for selecting the right variables 
make_lf_df <- function(x, master, lf_select, lf){
  temp <- lifeformMatch(x=x, master=master,
                           column1 = lf_select$column1[lf_select$name==lf][1], cat1=lf_select$cat1[lf_select$name==lf][1],
                           column2=lf_select$column2[lf_select$name==lf][1], cat2=lf_select$cat2[lf_select$name==lf],
                           name=lf)
  return(temp)
}

```
Tag raw data by lifeform
```{r}
tag_lf <- function(x, master, lf){
  
  df_temp <- list()
  for(i in 1:length(unique(lf$name))){
    temp_lf_name <- sort(unique(lf$name))[i]
      
    print(temp_lf_name)
    
    temp_lf <- subset(lf, name==temp_lf_name)
  
    df_lifeform <- master %>%
        filter((!!sym(temp_lf$column1[1])) %in% temp_lf$cat1[1]) %>%
        filter((!!sym(temp_lf$column2[1])) %in% temp_lf$cat2) %>%
        dplyr::select(size_class, aphia_id) %>%
        distinct()
    
    tf_lf <- data.frame(V1 = as.integer(paste(x$aphia_id, x$size_class) %in% paste(df_lifeform$aphia_id, df_lifeform$size_class)))
   
    names(tf_lf) <- temp_lf_name
    df_temp[[i]] <- tf_lf
  }
  sub_output <- do.call(cbind,df_temp)
  output <- cbind(x, sub_output)
  return(output)
}

df_abund_tagged <- tag_lf(x=df_abund, master=df_master, lf=lf_select)
```
Append taxon names for taxa with size-class specified
```{r}
df_abund_tagged$size_class_name[df_abund_tagged$sm_phyto == 1 & df_abund_tagged$size_class == 2] <- ",Sm"
df_abund_tagged$size_class_name[df_abund_tagged$lg_phyto == 1 & df_abund_tagged$size_class == 1] <- ",Lg"
df_abund_tagged$size_class_name[is.na(df_abund_tagged$size_class_name)] <- ""

df_abund_tagged$taxon <- paste0(df_abund_tagged$taxon, df_abund_tagged$size_class_name)

df_abund_tagged$size_class <- NULL
df_abund_tagged$size_class_name <- NULL
```
Aggregate the data by lifeform by sample
```{r}
count_lf <- function(x){
  data_id_list <- list()
  for(t in 1:length(unique(x$data_id))){
    
    data_id_temp <- sort(unique(x$data_id))[t]
    x_data_id <- as.data.frame(subset(x, data_id==data_id_temp))
    
    #isolate relevant lifeforms
    id_vars <- c("lon", "lat", "assess_id", "data_id", "contracting_party", "dataset_name", "year", "month", "day", "hour", "minute", "count", "taxon", "aphia_id")
  
    lifeforms <- sort(colnames(x_data_id)[!(colnames(x_data_id) %in% id_vars)])
    lifeforms <- colnames(x_data_id[lifeforms][colSums(x_data_id[lifeforms][lifeforms]) > 0])
    
    lf_list <- list()
    for(i in 1:length(unique(lifeforms))){
    
      df_lf_mon <- x_data_id %>%
                    dplyr::filter(get(lifeforms[i])==1) %>%
                    dplyr::group_by(data_id, lon, lat, year, month, day, hour, minute) %>%
                    dplyr::summarise(count=sum(count, na.rm=T),
                                     .groups = 'drop') %>%
                    dplyr::mutate(lifeform=lifeforms[i]) %>%
                    arrange(data_id, lifeform, year, month, day, hour, minute)

        
      lf_list[[i]] <- df_lf_mon
    }
    
    #spread and collapse to ensure proper inclusion of 0s
    df_data_id <- do.call(rbind, lf_list) %>%
      pivot_wider(names_from = lifeform, 
                  values_from = count,
                  values_fill = 0) %>%
      pivot_longer(-c("data_id", "year", "month", "day", "hour", "minute", "lat", "lon"), names_to = "lifeform", values_to = 'count')
    
    data_id_list[[t]] <- df_data_id
  }
  output <- do.call(rbind, data_id_list)
  return(output)
}

df_abund_agg <- count_lf(x=df_abund_tagged)
```
Replace 0 values with 0.5* the min value for the data_id/lifeform combination so data can be log transformed
```{r}
df_abund_agg <- df_abund_agg %>%
  group_by(data_id, lifeform) %>%
  dplyr::mutate(min_non_zero = min( count[count!=min(count)] )) %>%
  dplyr::mutate(count = count + min_non_zero*0.5) %>%
  ungroup() %>%
  dplyr::select(-min_non_zero)
```
Save the formatted data for ease of use rather than preparing the data each time it is needed
```{r}
#create data output directory
output_path <- paste0(dir_main, "Plankton_processed/")
dir.create(file.path(output_path), showWarnings = FALSE)

write_fst(df_abund_tagged, path=paste0(output_path, "plankton_by_taxon.fst"))
write_fst(df_abund_agg, path=paste0(output_path, "plankton_by_lifeform.fst"))
```





