---
title: "Integration of environmental drivers with plankton abundance time-series analysis"
---

This script is for loading the results of the modelling environmental drivers and integrating with Mann-Kendall statistic for the plankton lifeform abundance time-series. 

Author: Matthew Holland
Contact: matt.holland@plymouth.ac.uk
Latest version date: 17 December 2021
```{r}
#clear R environment
rm(list = ls()) 

#enter a rough bounding box for visualisation only (in decimal degrees format)
north <- 64
west <- -16
east <- 14
south <- 34

#reference period duration (in years)
ref_per <- 5

#enter the range of years covered by this analysis, the reference and comparison period will be calculated as the first n and the last n years of the dataset, with n determined by "ref_per" the variable
start_query <- 1960
end_query <- 2019

#enter the file directory path for the shapefile the data is partitioned with
path_shp_part <- "../Data_raw/COMP4_assessment_areas_v7e/"

#enter the filename of the shapefile the data is partitioned with
file_shp_part <- "COMP4_assessment_areas_v7e.shp"

#provide the name of the variable in the shapefile used for separating the assessment areas
assess_des <- "LongName"

#enter the main directory to use to access the processed data
dir_main <- "../Data_processed/"

#enter the directory where raw data is stored
dir_raw <- "../Data_raw/"

#enter the main directory to use to store image outputs
dir_out <- "../Output/"

#create plot output directory
output_path <- paste(dir_out, gsub(".shp", "", file_shp_part), "/", sep="")
dir.create(file.path(output_path), showWarnings = FALSE)

#generate directory for the plots
output_path_drivers <- paste(output_path, "drivers_", start_query, "_to_", end_query, "/", sep="")
dir.create(file.path(output_path_drivers), showWarnings = FALSE)
```
Load required packages and install if not already installed
```{r, include=FALSE}
#check if all required packages are installed. Install them if they are not present. 
#Then load required packages

#install rnaturalearthhires for map data
devtools::install_github("ropensci/rnaturalearthhires")

list.of.packages <- c("EnvStats", "dplyr", "ggplot2", "data.table", "tidyverse", "broom", "gridExtra", "pracma", "ggpattern", "fst", "sf", "lemon", "viridis", "RColorBrewer")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
rm(list.of.packages, new.packages)

options(scipen=999)

#switch off spherical geometry
sf::sf_use_s2(FALSE)
```
Generate string for the relevant subdirectory 
```{r}
#save the data as fst
dir_shp <- paste0(dir_main, gsub(".shp","",file_shp_part), "/")

#generate a string for the directory of the relevant CSV file
file_path <- paste0(dir_shp, gsub(".shp", "", file_shp_part), "_lifeforms", ".fst")

#load the Kendall results
df_fits_tot <- read_fst(path=paste0(gsub(".fst", "", file_path), "_kendall", ".fst"))
```
Import the results of the model fitting process from Step5 script
```{r}
#function for saving all models by dataset
load_rf_models <- function(dir){

  files <- sort(list.files(dir, full.names = T))
  
  df_list <- list()
  for(i in 1:length(files)){
    
    temp <- readRDS(files[i])
    
    df_list[[i]] <- temp
  }
  
  out <- do.call(rbind, df_list)
  return(out)
}
```
Integrate the model variable selections and the time-series analyses
```{r}
#generate directory for the plots
input_path_rf <- paste(dir_main, "RF_models", "/", sep="")

df_rf_orig <- load_rf_models(dir=input_path_rf)

df_rf <- df_rf_orig %>%
  dplyr::select(data_id, lifeform, assess_id, n_train, n_test, boruta_stats, rfboruta_rsquared) %>%
  relocate(boruta_stats, .after = last_col()) %>%
  dplyr::mutate(boruta_stats = map(boruta_stats, 
        ~ dplyr::mutate(.x, vars = rownames(.x)))
        ) %>%
  dplyr::mutate(boruta_stats = map(boruta_stats, ~ dplyr::select(.x, vars, meanImp, decision)))

boruta_stats <- df_rf %>%
  unnest_wider(boruta_stats) %>%
  group_by(data_id, lifeform, assess_id) %>%
  dplyr::summarise(across(c(vars, meanImp, decision), unlist), .groups = "drop") %>%
  dplyr::mutate(meanImp = round(meanImp, 4),
    meanImp = ifelse(decision=="Rejected",0,meanImp)) %>%
  dplyr::select(-decision) %>%
  pivot_wider(names_from=vars, values_from=meanImp, values_fill = NA) %>%
  dplyr::select(sort(tidyselect::peek_vars())) %>%
  relocate(data_id, lifeform, assess_id) %>%
  pivot_longer(-c(data_id, lifeform, assess_id), names_to = "param", values_to = "value", values_drop_na = TRUE) %>%
  dplyr::mutate(param = sub("\\_.*",'', param)) %>%
  pivot_wider(names_from = param, values_from = value, values_fill = NA)
```
Join all data into a combined dataframe
```{r}
df_comb <- df_rf %>%
  full_join(df_fits_tot) %>%
  dplyr::select(-c(n_test, boruta_stats)) %>%
  full_join(boruta_stats)
```
Select the dataset to display visually when there are multiple from different datasets and combine with the Kendall results
```{r, include=FALSE}
lifeforms <- c("diatom", "dinoflagellate", "holoplankton", "meroplankton", "gelatinous", "fishlarvae", "lg_copepods", "sm_copepods")
lifeforms_rep <- c("Diatoms", "Dinoflagellates", "Holoplankton", "Meroplankton", "Gelatinous", "Fish larvae/eggs", "Large copepods", "Small copepods")

#subset to lifeforms of interest for the final assessment
df_summary <- df_comb %>%
  group_by(assess_id, lifeform) %>% #determine the datasets with the most months of real samples in an assessment unit
  filter(unique_months == max(unique_months, na.rm=T)) %>%
  ungroup() %>%
  dplyr::select(-unique_months, -prop_years)

df_unique_months <- df_summary %>%
  pivot_longer(cols = -c(1:8), names_to = "param", values_to = "imp") %>%
  filter(!is.na(imp)) %>%
  arrange(assess_id, lifeform, -imp) %>%
  filter(param!="year") %>%
  dplyr::mutate(param = factor(param, levels=sort(unique(param)))) %>%
  dplyr::mutate(param_type = recode(param,
                        sst="Temp",
                        wspd="Wind",
                        precip="Precip",
                        mld="MLD",
                        cvel="Current",
                        nppv="Productivity",
                        attn="Attenuation",
                        psal="Salinity",
                        amo="Climate_index",
                        nao="Climate_index",
                        phos="Nutrients",
                        ptot="Nutrients",
                        slca="Nutrients",
                        ntot="Nutrients",
                        ntra="Nutrients",
                        amon="Nutrients",
                        year="Year",
                        spco2="Carbonate")) %>%
  group_by(assess_id, lifeform) %>%
  filter(imp==max(imp, na.rm=T)) %>%
  ungroup() %>%
  dplyr::mutate(param_type = factor(param_type, levels=sort(as.character(unique(param_type)))))

```
Load map data
```{r}
#load shapefile of European landmass for display
coast <- rnaturalearth::ne_countries(returnclass = "sf", scale = "large")

#load the shapefile associated with the data
shp_part <- st_read(paste(path_shp_part, file_shp_part, sep=""))

#choose the variable to use as the primary separator of assessment areas
shp_part$assess_id <- shp_part[[assess_des]]
```
Create sf with multi-geometry (polygon and point)
```{r}
#import point coordinates of datasets
df_datasets <- fread(file=paste0(dir_main, "dataset_ids_and_point_id.csv"))

assess_id_list <- df_datasets %>%
    filter(is_point) %>%
    dplyr::select(assess_id, mean_lon, mean_lat) %>%
    distinct() %>%
    filter(unique(assess_id) %in% all_of(unique(df_unique_months$assess_id))) %>%
    dplyr::rename("x"="mean_lon",
           "y"="mean_lat")

pts <- st_as_sf(assess_id_list, coords = c("x","y"), remove = FALSE)
st_crs(pts) <- 4326
pts <- st_transform(pts, crs=st_crs(shp_part))
pts <- pts %>% dplyr::select(assess_id, geometry) %>% dplyr::mutate(is_point=TRUE, is_plume=FALSE)
```
Create separate designation for river plumes
```{r, include=FALSE}
#create sf object for points
plume_list <- shp_part %>%
    dplyr::select(assess_id, geometry) %>%
    distinct() %>%
    filter(grepl("plume",tolower(assess_id))) %>%
    dplyr::mutate(geometry = st_transform(geometry, 2163)) %>%
    dplyr::mutate(coords = st_centroid(geometry)) %>%
    dplyr::mutate(geometry = NULL) %>%
    dplyr::mutate(coords = st_transform(coords, 4326)) %>%
    unnest_wider(coords) %>%
    dplyr::rename("x"="...1",
         "y"="...2")

plumes <- st_as_sf(plume_list, coords = c("x","y"), remove = FALSE)
st_crs(plumes) <- 4326
plumes <- st_transform(plumes, crs=st_crs(shp_part))
plumes <- plumes %>% dplyr::select(assess_id, geometry) %>% dplyr::mutate(is_point=FALSE, is_plume=TRUE) %>% dplyr::select(names(pts))

```
Merge the points and plumes coordinates with the shapefile
```{r}
#merge geometry of points and polygons
shp_comb <- shp_part %>%
  dplyr::mutate(is_point=FALSE, is_plume=FALSE) %>%
  dplyr::select(names(pts)) %>%
  bind_rows(pts) %>%
  bind_rows(plumes)
```
Combine the variable importance data with the map data
```{r}
lf_shp <- expand.grid(lifeform = unique(df_unique_months$lifeform), assess_id = unique(shp_comb$assess_id)) %>%
  arrange(lifeform, assess_id)

shp_merged <- merge(shp_comb, lf_shp, by="assess_id", all=T)
shp_merged <- merge(shp_merged, df_unique_months, by=c("assess_id", "lifeform"), all.x=TRUE)

st_crs(shp_merged) <- 4326

#arrange factor levels 
shp_merged$param_type <- factor(shp_merged$param_type, levels=sort(unique(shp_merged$param_type)))
shp_merged$assess_id <- factor(shp_merged$assess_id, levels=sort(unique(shp_merged$assess_id)))
```
Summarise in map form
```{r}
#rsquared threshold
rsq_thr <- 0.2

#importance threshold
imp_thr <- 0

#generate colour palette
my.cols <- brewer.pal(length(levels(shp_merged$param_type)), "Set3")
my.cols[my.cols=="#D9D9D9"] <- "#FFED6F" #replace grey in palette

my.cols[6] <- "#B3DE69"
my.cols[7] <- "#80B1D3"
my.cols[5] <- "#FDB462"

#generate the plot output
  ggplot()+
  geom_sf(data=coast, inherit.aes=F, fill="grey80", colour="grey40", lwd = 0.2)+
  geom_sf(data=subset(shp_merged, 
                      is_point == F & 
                        is_plume==F), 
          fill="white", colour="black", lwd = 0.2)+
    geom_sf(data=subset(shp_merged, 
                      is_point == F & 
                        is_plume==F &
                        is.na(statistic)), 
          fill="white", colour="black", lwd = 0.2)+
  geom_sf(data=subset(shp_merged, 
                      (is_point == F & 
                        is_plume == F &
                        sig == T) &
                        (rfboruta_rsquared < rsq_thr |
                        imp < imp_thr)), 
          fill="grey60", colour="black", lwd = 0.2)+
  geom_sf(data=subset(shp_merged, 
                      is_point == F &
                        is_plume == F &
                         sig == T &
                        !is.na(imp) &
                        rfboruta_rsquared >= rsq_thr &
                        imp >= imp_thr), 
          aes(fill=param_type), colour="black", lwd = 0.1, alpha=0.9)+
  geom_sf(data=subset(shp_merged, 
                        is_plume == T &
                        sig == T &
                        !is.na(imp) & 
                        rfboruta_rsquared >= rsq_thr &
                        imp >= imp_thr), 
          aes(fill=param_type), colour="black", alpha=0.9, shape=24, size=3)+
  geom_sf(data=subset(shp_merged, 
                        is_point == T &
                        sig == T &
                        !is.na(imp) & 
                        rfboruta_rsquared >= rsq_thr &
                        imp >= imp_thr), 
          aes(fill=param_type), colour="black", alpha=0.9, shape=21, size=3)+
  facet_rep_wrap(~lifeform, repeat.tick.labels = TRUE, ncol=4)+
  coord_sf(xlim=c(west, east), ylim=c(south, north), expand=FALSE)+
  scale_fill_manual("Most\nimportant\npressure", values=my.cols)+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust=1, vjust=1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
      strip.text.x = element_text(size = 9))+
  guides(fill=guide_legend(override.aes=list(shape=NA)))
  
  ggsave(file=paste(output_path_drivers, "lifeforms", ".png", sep=""), 
         height=40, width=30, units="cm", bg="white", dpi=500)


```






